---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

### Load packages and set working directory
```{r loaddata, echo=TRUE, results='hide'}
library(dplyr)
library(ggplot2)
setwd("~/prof-dev/JHU-Rprogramming/Reproducible-Research/week2")
data <- read.csv("./data/activity.csv")
```

### Convert date to date class
```{r getdate, echo=TRUE}
data$date <- as.Date(strptime(data$date, format="%Y-%m-%d"))
```

## Total number of steps per day
```{r dailysteptotal, echo=TRUE}
dailyTotal <- aggregate(data$steps, by=list(data$date),
                            FUN=sum, na.rm=TRUE)
colnames(dailyTotal) <- c("date", "steps")
```

Here is a histogram of the total number of steps taken each day
```{r dailystephist, fig.height=4}
par(mar = c(5,4,1,1), las=1)
hist(dailyTotal$steps, breaks = 20,
        main = "Total steps per day", 
        ylab = "steps",
        xlab = "")

```

### What is mean total number of steps taken per day?
```{r mean, echo=TRUE}
as.integer(mean(dailyTotal$steps, na.rm=TRUE))
```

### What is median total number of steps taken per day?
```{r median, echo=TRUE}
median(dailyTotal$steps, na.rm=TRUE)
```

## What is the average daily activity pattern? 
Here is a plot:
```{r dailyactivity, fig.height=4}
dailyActivity <- aggregate(data$steps, by=list(data$interval),
                            FUN=mean, na.rm=TRUE)
colnames(dailyActivity) <- c("interval", "steps")
plot(dailyActivity$interval,dailyActivity$steps,
     type = "l",
     xlab = "5-min interval",
     ylab = "mean no. of steps")
```

```{r busyTime, echo=TRUE}
busyTime <- dailyActivity$interval[which.max(dailyActivity$steps)]
```
The interval with the most steps is `r busyTime` 

## Imputing missing values
There are `r sum(is.na(data))` missing values in the data.

Where steps is NA, replace with the median for that 5-minute interval:

```{r impute, echo=TRUE}
#impute.median <- function(x) replace(x, is.na(x), median(x, na.rm = TRUE))
#data %>%
#  group_by(interval) %>%
#  mutate(steps = impute.median(steps))

data %>% 
  group_by(interval) %>% 
  mutate(steps = ifelse(is.na(steps), median(steps, na.rm = TRUE), steps))
```

After updating the daily total number of steps with the imputed data, here is a histogram of the total number of steps taken each day
```{r dailystephist2, fig.height=4}
par(mar = c(5,4,1,1), las=1)
hist(dailyTotal$steps, breaks = 20,
        main = "Total steps per day", 
        ylab = "steps",
        xlab = "")

```

The mean total number of steps taken per day is `r as.integer(mean(dailyTotal$steps))`

And the median total number of steps taken per day is `r median(dailyTotal$steps)`

My method of imputation has not led to significant changes in the patterns observed before imputation.

## Are there differences in activity patterns between weekdays and weekends?

```{r weekday, echo=TRUE}
data$day <- weekdays(data$date)
data$wkday <- factor(ifelse((data$day=="Saturday" |data$day=="Sunday"), "weekend", "weekday"))

wkdayActivity <- group_by(data, interval, wkday) %>% 
  summarize(steps = mean(steps, na.rm = TRUE))
```

Here is a plot of steps by weekday vs weekend:
```{r wkdayplot, fig.height=4, echo=TRUE}
ggplot(data=wkdayActivity, aes(x=interval, y=steps)) +
  geom_bar(colour="black", stat="identity") +
  facet_wrap(~wkday, nrow=2, ncol=1) +
  ggtitle("Steps by Weekend vs Weekday")

```

We see a peak in activity in the mornings during weekdays, while activity is more evenly spread throughout the day on weekends.